Statically Typed String Sanitation Inside a Python (Technical Report)

### 2014-12-14 sanitation-psp14/tr/strings_chapter.tex
prog.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{thm}[Safety for $\lambdap$] If $\emptyset \vdash \iota : \tau$ 
  then $\iota \sreduces \dot v$ and $\emptyset \vdash \dot v : \tau$.
\end{thm}

@@

\begin{mathpar}
\inferrule[SP-E-Abs]
{ \ }
{\lambda x.e \sreduces \lambda x.e}

\inferrule[SP-E-App]
{ \iota_1 \sreduces \lambda x . \iota_3 \\  \iota_2 \sreduces \dot{v}_2 \\ [\dot{v}_2 / x] \iota_3 \sreduces \dot{v}_3}
{ \iota_1(\iota_2) \sreduces \dot{v}_3}

\inferrule[SP-E-Str]
{ \ }
{\tstr{s} \treduces \tstr{s}}

\inferrule[SP-E-Rx]
{ \ }
{\rx{r} \treduces \rx{r}}

\inferrule[SP-E-Concat]
{\iota_1 \treduces \tstr{s_1} \\ \iota_2 \treduces \tstr{s_2}} 
{\tconcat{\iota_1}{\iota_2} \treduces \tstr{s_1 s_2}} % ???

\inferrule[SP-E-Case-$\epsilon$]
{
  \iota_1 \treduces \tstr{ \epsilon } \\
  \iota_2 \treduces \dot{v_2} \\
}
{
  \pstrcase{\iota_1}{\iota_2}{x,y.\iota_3} \treduces \dot{v_2}
}

\inferrule[SP-E-Case-Concat]
{
  \iota_1 \treduces \tstr{as} \\
  [\tstr{a},\tstr{s}/x,y] \iota_3 \treduces \dot{v} 
}
{
  \pstrcase{\iota_1}{\iota_2}{x,y.\iota_3} \treduces \dot{v}
}

\inferrule[SP-E-Replace]
{\iota_1 \treduces \rx{r} \\ \iota_2 \treduces \tstr{s_2} \\ \iota_3 \treduces \tstr{s_3} \\ \lsubst{r}{s_2}{s_3} = s} 
{\preplace{\iota_1}{\iota_2}{\iota_3} \treduces \tstr{s}}

\inferrule[SP-E-Check-OK]
{
  \iota_r \treduces \rx{r} \\
  \iota \treduces \tstr{s} \\ 
s \in \lang{r} \\ 
\iota_1 \treduces \dot{v_1} 
}
{
  \tcheck{\iota_r}{\iota}{\iota_1}{\iota_2} \treduces \dot{v_1}
}

\inferrule[SP-E-Check-NotOK]
{\iota_r \treduces \rx{r} \\ \iota \treduces \tstr{s} \\ s \not \in \lang{r} \\ \iota_2 \treduces \dot{v_2}}
{\tcheck{\iota_r}{\iota}{\iota_1}{\iota_2} \treduces \dot{v_2}}
\end{mathpar}
\caption{Small step semantics for $\lambdap$}.
\label{fig:sdlambdap}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% LAMBDAP SECTION

@@

This section follows the same structure as the safety proof for $\lambdas$ --
we prove type safety for a small-step semantics, prove a semantic correspondence,
and then transfer the safety result to the big-step semantics in the paper.

Our proofs here are more sketchy than in the $\lambdas$ section both because the
proof method is already demonstrated, and becaues $\lambdap$ is a fairly straightforward
extension of $\lambda$ for which these results are already estbalished.

\begin{thm}[Progress]
If $\emptyset\vdash \iota : \tau$ either $\iota = \dot{v}$ or $\iota \mapsto \iota'$ for some $\iota'$.
\end{thm}
\begin{proof}
The proof proceeds by induction on the typing assumption. We elide contexts where context has obviously not changed.

Consider only the string and regex (non-$\lambda$) fragments of $\lambdap$.

Regular expressions are always values.

For strings, the concatenation cases are trivial.

\item \textbf{P-T-Case}.
Suppose $\emptyset \vdash \pstrcase{\iota_1}{\iota_2}{x,y.\iota_3}$.
By inversion, $\iota_1 : \str$ and so either $\iota_1 \mapsto \iota_1'$
or by canonical forms, $\iota_1 = \tstr{s_1}$.
Similarly, $\iota_2 \mapsto \iota_2'$ or else $\iota_2 = \tstr{s_2}$.
In the former cases, progress occurs via the compatibility rules.
in the case where both are string values, progress occurs via the case concatenation rule.

\item \textbf{P-T-Replace}.
Suppose $\emptyset \vdash \preplace{\iota_1}{\iota_2}{\iota_3}$.
By inversion, $\iota_1 : \regex$ and so by canonical forms $\iota_1 = \rx{r}$.
By inversion, $\iota_2 : \str$ and so by induction either $\iota_2 \mapsto \iota_2'$
or else $\iota_2 = \tstr{s_2}$ for some string $s_2$.
Similarly, either $\iota_3$ steps or else $\iota_3 = \tstr{s_3}$.
In case any steps occur, progress occurs. In the remaining case,
PP-E-Replace applies and so progress occurs.

\item \textbf{P-T-Check}.
Finally, suppose $\emptyset \vdash \check{\iota_x}{\iota_1}{\iota_2}{\iota_3}$.
In case any of these step, then progress occurs.
In the remaining cases, applications of inversion and canonical forms for each
$\iota_x$ and $\iota_1$ implies that the term at hand equals $\check{\rx{r}}{\tstr{s}}{\iota_2}{\iota_3}$,
which evaluates to either $\iota_2$ or $\iota_3$.

\begin{thm}[Preservation]
If $\emptyset \vdash \iota : \tau$ and $\iota \mapsto \iota'$ then $\emptyset \vdash \iota' : \tau$.
\end{thm}
\begin{proof}
\todo{Fill in if there's time. Otherwise there's nothing interesting going on.}
\end{proof}

\begin{thm}[Safety for Small-Step Semantics]
If $\iota \sreduces \dot{v}$ and $\iota \mapsto \iota'$ then $\iota' \sreduces \dot{v}$. 
\end{thm}
\begin{proof}
A direct result from progress and preservation.
\end{proof}

We only prove semantic correspondence in one direction; again, whereas $\lambdas$ proofs
were detailed, here we provide a less verbosy proof.

\begin{thm}[Semantic Correspondence]
If $\iots \mapsto^* \iota'$ then $\iota \sreduces \iota'$.
\end{thm}
\begin{proof}
By induction on the structure of $\iota$.

\begin{itemize}[label=$ $,itemsep=1ex]
\end{itemize}
\end{proof}

\begin{thm}[Safety for $\lambdap$] If $\emptyset \vdash \iota : \tau$ 
  then $\iota \sreduces \dot v$ and $\emptyset \vdash \dot v : \tau$.
\end{thm}
\begin{proof}
If $\emptyset \vdash \iota : \tau$ then $\iota \mapsto^* \iota'$.
Therefore, $\iota \sreduces \iota'$ by part 2 of the semantic correspondence theorem.

Since $\emptyset \vdash \iota : \tau$ and $\iota \mapsto^* \iota'$, it follows that $\emptyset \vdash \iota' : \tau$ by type safety for the small step semantics.

@@

\begin{mathpar}
\inferrule[PP-E-ConcatLeft]
{\iota_1 \mapsto \iota_1'}
{\tconcat{\iota_1}{\iota_2} \treduces \tconcat{\iota_1'}{\iota_2}}

\inferrule[PP-E-ConcatRight]
{\iota_2 \mapsto \iota_2'}
{\tconcat{\iota_1}{\iota_2} \treduces \tconcat{\iota_1}{\iota_2'}}

\inferrule[PP-E-Concat]
{\ }
{\tconcat{\tstr{s_1}}{\tstr{s_2}} \treduces \tstr{s_1s_2}}

\inferrule[PP-E-CaseLeft]
{ \iota_1 \mapsto \iota_1' }
{ \pstrcase{\iota_1}{\iota_2}{x,y.\iota_3} \mapsto \pstrcase{\iota_1'}{\iota_2}{x,y.\iota_3} }

\inferrule[PP-E-CaseRight]
{ \iota_2 \mapsto \iota_2' }
{ \pstrcase{\dot{v}}{\iota_2}{x,y.\iota_3} \mapsto \pstrcase{\dot{v}}{\iota_2'}{x,y.\iota_3} }

\inferrule[PP-E-Case]
{ }
{ \pstrcase{\tstr{s_1}}{\tstr{s_2}}{x,y.\iota_3} \mapsto \tstr{s_1s_2}}

\inferrule[PP-E-ReplaceLeft]
{\iota_1 \mapsto \iota_1'}
{ \preplace{\iota_1}{\iota_2}{\iota_3} \mapsto \preplace{\iota_1'}{\iota_2}{\iota_3} }

\inferrule[PP-E-ReplaceMid]
{\iota_2 \mapsto \iota_2'}
{ \preplace{\rx{r}}{\iota_2}{\iota_3} \mapsto \preplace{\rx{r}}{\iota_2'}{\iota_3} }

\inferrule[PP-E-ReplaceRight]
{\iota_3 \mapsto \iota_3'}
{ \preplace{\rx{r}}{\tstr{s_2}}{\iota_3} \mapsto \preplace{\rx{r}}{\tstr{s_2}}{\iota_3'} }

\inferrule[PP-E-Replace]
{\ }
{ \preplace{\rx{r}}{\tstr{s_2}}{\tstr{s_3}} \mapsto \tstr{\lsubst{r}{s_2}{s_3}} }

\inferrule[PP-E-Check-Left]
{ \iota_x \mapsto \iota_x' }
{ \check{\iota_x}{\iota}{\iota_1}{\iota_2} \mapsto \check{\iota_x}{\iota}{\iota_1}{\iota_2} }

\inferrule[PP-E-Check-Right]
{ \iota \mapsto \iota' }
{ \check{\rx{r}}{\iota}{\iota_1}{\iota_2} \mapsto \check{\rx{r}}{\iota'}{\iota_1}{\iota_2} }

\inferrule[PP-E-Check-OK]
{ s \in \lang{r} }
{ \check{\rx{r}}{\tstr{s}}{\iota_1}{\iota_2} \mapsto \iota_1}

\inferrule[PP-E-Check-OK]
{ s \not\in \lang{r} }
{ \check{\rx{r}}{\tstr{s}}{\iota_1}{\iota_2} \mapsto \iota_2}

\end{mathpar}
\caption{Small step semantics for $\lambdap$ (extends L-E rules)}.
\label{fig:sdlambdap}
\end{figure}
